---
title: "Robust BPST Simulation"
author: "Yang Long"
date: "2023-04-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(Triangulation)
library(BPST)
library(tidyverse)
library(plotly)
library(pense)
library(robustbase)
source("robust_bpst_functions.R")
set.seed(20230426)
```

In this section, we perform simulation study of robust BPST using the setup in Example 1 of Lai and Wang (2013). 
Specifically, for the regression problem
$$
Y=\mu(x_1,x_2) + \epsilon,
$$
we consider the following two signal functions: 

- Linear
$$
\mu_1(x_1, x_2) = 10x_1 + x_2 + 19
$$

- Sinusoidal: 
$$
\mu_2(x_1,x_2) = 24 + 5\sin\left(\pi(x_1^2+x_2^2)\right)
$$

### Data Generation 

a $101 \times 101$ grid with values evenly spaced between 0 and 1 is created on a square domain 
$[0,1]^2$

```{r echo=TRUE}
# create boundary
bb <- rbind(c(0, 0), c(1, 0), c(1, 1), c(0, 1), c(0, 0))

# grid resolution
grid_size <- c(101, 101)

# create coordinates
xy_grid <- expand.grid(seq(min(bb[,1]), max(bb[,1]), length = grid_size[1]), 
                       seq(min(bb[,2]), max(bb[,2]), length = grid_size[2]))
Z <- as.matrix(xy_grid)

# True signal 
true_sig_1 <- apply(Z, 1, function(x) 10*x[1] + x[2] + 19)
true_sig_2 <- apply(Z, 1, function(x) 24 + 5*sin(pi*(x[1]^2 + x[2]^2)))

# generate a sample of 2000 observations and different levels of contamination 
sample_size <- 2000

generate_sample <- function(resp_all, grid_size, sample_size, contamination) {
  selected <- sample(prod(grid_size), size = sample_size)
  resp_sample <- resp_all[selected]
  cont_ind <- sample(sample_size, size = sample_size * contamination)
  noise <- rep(0, sample_size)
  noise[cont_ind] <- rnorm(length(cont_ind), mean = 20, sd = 3)
  noise[-cont_ind] <- rnorm(sample_size - length(cont_ind), mean = 0, sd = 1)
  resp_sample <- resp_sample + noise 
  
  list(
    index = selected, 
    grid_samples = Z[selected, ],
    Y = resp_sample
  )
}

# we generate two samples, one with 20% contamination, and the other one with 40% contamination 
sample_1_1 <- generate_sample(true_sig_1, grid_size = grid_size, sample_size = sample_size, contamination = 0.2)
sample_1_2 <- generate_sample(true_sig_1, grid_size = grid_size, sample_size = sample_size, contamination = 0.4)

Y1_1 <- sample_1_1$Y
Y1_2 <- sample_1_2$Y

# same for the second setup 
sample_2_1 <- generate_sample(true_sig_2, grid_size = grid_size, sample_size = sample_size, contamination = 0.2)
sample_2_2 <- generate_sample(true_sig_2, grid_size = grid_size, sample_size = sample_size, contamination = 0.4)

Y2_1 <- sample_2_1$Y
Y2_2 <- sample_2_2$Y
```

### Triangulation 

Next we triangulate the domain using `TriMesh()` function from the `Triangulation` package.

```{r warning=FALSE}
# triangulation
VT <- TriMesh(bb, n = 3)
# vertices 
V <- VT$V
# triangles
Tr <- VT$Tr
```

### LS fitting 

```{r warning=FALSE}
deg <- 3

mfit1_1 <- fit.BPST(Y = Y1_1, Z = sample_1_1$grid_samples, V = V, Tr = Tr, d = deg)
mfit1_2 <- fit.BPST(Y = Y1_2, Z = sample_1_2$grid_samples, V = V, Tr = Tr, d = deg)
mfit2_1 <- fit.BPST(Y = Y2_1, Z = sample_2_1$grid_samples, V = V, Tr = Tr, d = deg)
mfit2_2 <- fit.BPST(Y = Y2_2, Z = sample_2_2$grid_samples, V = V, Tr = Tr, d = deg)

pred1_1 <- predict.BPST(mfit1_1, Zpred = Z)
pred1_2 <- predict.BPST(mfit1_2, Zpred = Z)
pred2_1 <- predict.BPST(mfit2_1, Zpred = Z)
pred2_2 <- predict.BPST(mfit2_2, Zpred = Z)

z1_1 <- matrix(pred1_1$Ypred, nrow = 101, byrow = TRUE)
z1_2 <- matrix(pred1_2$Ypred, nrow = 101, byrow = TRUE)
z2_1 <- matrix(pred2_1$Ypred, nrow = 101, byrow = TRUE)
z2_2 <- matrix(pred2_2$Ypred, nrow = 101, byrow = TRUE)

df1_1 <- data.frame(sample_1_1$grid_samples, Y1_1)
df1_2 <- data.frame(sample_1_2$grid_samples, Y1_2)
df2_1 <- data.frame(sample_2_1$grid_samples, Y2_1)
df2_2 <- data.frame(sample_2_2$grid_samples, Y2_2)
```

```{r fig.height = 4, fig.width = 10, fig.align = "center", out.width='100%', warning=FALSE}
plotly_figure1_1 <- plot_ly(
  x = unique(xy_grid[, 1]),
  y = unique(xy_grid[, 2]),
  z = z1_1,
  type = "surface",
  scene = 'scene1'
) %>%
  # add_surface(x = unique(xy_grid[, 1]),
  #             y = unique(xy_grid[, 2]),
  #             z = matrix(true_sig_1, nrow = 101, byrow = TRUE)) %>%
  add_trace(data = df1_1, x = df1_1[ , 1], y = df1_1[ , 2], z = df1_1[ , 3],
            mode = "markers", type = "scatter3d",
            marker = list(size = 1, color = "black")) %>%
  layout(scene1 = list(camera = list(
    eye = list(
      x = -1.25,
      y = -1.25,
      z = 1.25
    ),
    center = list(x = 0,
                  y = 0,
                  z = 0)
  )))

plotly_figure1_2 <- plot_ly(
  x = unique(xy_grid[, 1]),
  y = unique(xy_grid[, 2]),
  z = z1_2,
  type = "surface",
  scene = 'scene2'
) %>%
  # add_surface(x = unique(xy_grid[, 1]),
  #             y = unique(xy_grid[, 2]),
  #             z = matrix(true_sig_1, nrow = 101, byrow = TRUE)) %>%
  add_trace(data = df1_2, x = df1_2[ , 1], y = df1_2[ , 2], z = df1_2[ , 3],
            mode = "markers", type = "scatter3d",
            marker = list(size = 1, color = "black")) %>%
  layout(scene2 = list(camera = list(
    eye = list(
      x = -1.25,
      y = -1.25,
      z = 1.25
    ),
    center = list(x = 0,
                  y = 0,
                  z = 0)
  )))

plotly_figure2_1 <- plot_ly(
  x = unique(xy_grid[, 1]),
  y = unique(xy_grid[, 2]),
  z = z2_1,
  type = "surface",
  scene = 'scene3'
) %>%
  # add_surface(x = unique(xy_grid[, 1]),
  #             y = unique(xy_grid[, 2]),
  #             z = matrix(true_sig_1, nrow = 101, byrow = TRUE)) %>%
  add_trace(data = df2_1, x = df2_1[ , 1], y = df2_1[ , 2], z = df2_1[ , 3],
            mode = "markers", type = "scatter3d",
            marker = list(size = 1, color = "black")) %>%
  layout(scene3 = list(camera = list(
    eye = list(
      x = -1.25,
      y = -1.25,
      z = 1.25
    ),
    center = list(x = 0,
                  y = 0,
                  z = 0)
  )))

plotly_figure2_2 <- plot_ly(
  x = unique(xy_grid[, 1]),
  y = unique(xy_grid[, 2]),
  z = z2_2,
  type = "surface",
  scene = 'scene4'
) %>%
  # add_surface(x = unique(xy_grid[, 1]),
  #             y = unique(xy_grid[, 2]),
  #             z = matrix(true_sig_1, nrow = 101, byrow = TRUE)) %>%
  add_trace(data = df2_2, x = df2_2[ , 1], y = df2_2[ , 2], z = df2_2[ , 3],
            mode = "markers", type = "scatter3d",
            marker = list(size = 1, color = "black")) %>%
  layout(scene4 = list(camera = list(
    eye = list(
      x = -1.25,
      y = -1.25,
      z = 1.25
    ),
    center = list(x = 0,
                  y = 0,
                  z = 0)
  )))

subplot(plotly_figure1_1, plotly_figure1_2, plotly_figure2_1, plotly_figure2_2, nrows=2)
```

### BPST S-estimator

```{r}
# use a random subsample to create an initial solution 
generate_initial <- function(B, Y, lamb, sample_size, subsample_size, num_init = 5) {
  solution <- lapply(1:num_init, function(i) {
    sub_sample <- sample(sample_size, subsample_size)
    B_sub <- B[sub_sample,]
    Y_sub <- Y[sub_sample]
    solve(t(Q2) %*% (t(B_sub)%*%B_sub + lamb * K) %*% Q2) %*% t(Q2) %*% t(B_sub) %*% Y_sub
  })
  
  do.call(cbind, solution)
}
```

```{r}
# for first setup 
lamb <- 1e6
subsample_size <- 500
num_init <- 3


# for 20% contamination 
Bfull <- basis(V,Tr,d=deg,r=1,Z=sample_1_1$grid_samples,Hmtx=TRUE,Kmtx=TRUE,QR=TRUE,TA=TRUE)
B <- Bfull$B
H <- Bfull$H
Q2 <- Bfull$Q2
K <- Bfull$K
gamma <- matrix(0, nrow = dim(B)[2], ncol = num_init)
obj_fun_val <- rep(0, num_init)
iter1_1 <- rep(0, num_init)

theta_start_1_1 <- generate_initial(B, Y1_1, lamb, sample_size, subsample_size, num_init = num_init)
for (i in 1:num_init) {
  sfit_1 <- robust.BPST.est.s.1(B, Q2, K, lambda = lamb, Y = Y1_1, 
                                rho = "bisquare", bdp = 0.5, 
                                theta_start = as.matrix(theta_start_1_1[,i]),
                                maxit = 500, tol = 1e-6)
  iter1_1[i] <- sfit_1$iteration
  gamma[,i] <- drop(sfit_1$gamma)
  obj_fun_val[i] <- sfit_1$obj_fun
}
gamma_s_1_1 <- gamma[, which.min(obj_fun_val)]

# for 40% contamination 
Bfull <- basis(V,Tr,d=deg,r=1,Z=sample_1_2$grid_samples,Hmtx=TRUE,Kmtx=TRUE,QR=TRUE,TA=TRUE)
B <- Bfull$B
H <- Bfull$H
Q2 <- Bfull$Q2
K <- Bfull$K
gamma <- matrix(0, nrow = dim(B)[2], ncol = num_init)
obj_fun_val <- rep(0, num_init)
iter1_2 <- rep(0, num_init)

theta_start_1_2 <- generate_initial(B, Y1_2, lamb, sample_size, subsample_size, num_init = num_init)
for (i in 1:num_init) {
  sfit_2 <- robust.BPST.est.s.1(B, Q2, K, lambda = lamb, Y = Y1_2, 
                                rho = "bisquare", bdp = 0.5, 
                                theta_start = as.matrix(theta_start_1_2[,i]),
                                maxit = 500, tol = 1e-6)
  iter1_2[i] <- sfit_2$iteration
  gamma[,i] <- drop(sfit_2$gamma)
  obj_fun_val[i] <- sfit_2$obj_fun
}
gamma_s_1_2 <- gamma[, which.min(obj_fun_val)]
```

```{r}
# for second setup 
lamb <- 0.00001
# for 20% contamination 
Bfull <- basis(V,Tr,d=deg,r=1,Z=sample_2_1$grid_samples,Hmtx=TRUE,Kmtx=TRUE,QR=TRUE,TA=TRUE)
B <- Bfull$B
H <- Bfull$H
Q2 <- Bfull$Q2
K <- Bfull$K
gamma <- matrix(0, nrow = dim(B)[2], ncol = num_init)
obj_fun_val <- rep(0, num_init)
iter2_1 <- rep(0, num_init)

theta_start_2_1 <- generate_initial(B, Y2_1, lamb, sample_size, subsample_size, num_init = num_init)
for (i in 1:num_init) {
  sfit_1 <- robust.BPST.est.s.1(B, Q2, K, lambda = lamb, Y = Y2_1, 
                                rho = "bisquare", bdp = 0.5, 
                                theta_start = as.matrix(theta_start_2_1[,i]),
                                maxit = 500, tol = 1e-6)
  iter2_1[i] <- sfit_1$iteration
  gamma[,i] <- drop(sfit_1$gamma)
  obj_fun_val[i] <- sfit_1$obj_fun
}
gamma_s_2_1 <- gamma[, which.min(obj_fun_val)]

# for 40% contamination 
Bfull <- basis(V,Tr,d=deg,r=1,Z=sample_2_2$grid_samples,Hmtx=TRUE,Kmtx=TRUE,QR=TRUE,TA=TRUE)
B <- Bfull$B
H <- Bfull$H
Q2 <- Bfull$Q2
K <- Bfull$K
gamma <- matrix(0, nrow = dim(B)[2], ncol = num_init)
obj_fun_val <- rep(0, num_init)
iter2_2 <- rep(0, num_init)

theta_start_2_2 <- generate_initial(B, Y2_2, lamb, sample_size, subsample_size, num_init = num_init)
for (i in 1:num_init) {
  sfit_2 <- robust.BPST.est.s.1(B, Q2, K, lambda = lamb, Y = Y2_2, 
                                rho = "bisquare", bdp = 0.5, 
                                theta_start = as.matrix(theta_start_2_2[,i]),
                                maxit = 500, tol = 1e-6)
  iter2_2[i] <- sfit_2$iteration
  gamma[,i] <- drop(sfit_2$gamma)
  obj_fun_val[i] <- sfit_2$obj_fun
}
gamma_s_2_2 <- gamma[, which.min(obj_fun_val)]
```

```{r fig.height = 4, fig.width = 10, fig.align = "center", out.width='100%', warning=FALSE}

B_pred <- basis(V,Tr,d=deg,r=1,Z,FALSE,FALSE,FALSE,FALSE)
zs1_1 <- B_pred$B %*% as.matrix(gamma_s_1_1)
zs1_2 <- B_pred$B %*% as.matrix(gamma_s_1_2)
zs2_1 <- B_pred$B %*% as.matrix(gamma_s_2_1)
zs2_2 <- B_pred$B %*% as.matrix(gamma_s_2_2)


plotly_figure_s_1_1 <- plot_ly(
  x = unique(xy_grid[, 1]),
  y = unique(xy_grid[, 2]),
  z = z1_1,
  type = "surface", 
  scene = "scene1"
) %>%
  add_surface(x = unique(xy_grid[, 1]),
              y = unique(xy_grid[, 2]),
              z = matrix(zs1_1, nrow = 101, byrow = TRUE)) %>% 
  add_trace(data = df1_1, x = df1_1[ , 1], y = df1_1[ , 2], z = df1_1[ , 3],
            mode = "markers", type = "scatter3d",
            marker = list(size = 1, color = "black"))%>%
  layout(scene5 = list(camera = list(
    eye = list(
      x = -1.25,
      y = -1.25,
      z = 1.25
    ),
    center = list(x = 0,
                  y = 0,
                  z = 0)
  )))

plotly_figure_s_1_2 <- plot_ly(
  x = unique(xy_grid[, 1]),
  y = unique(xy_grid[, 2]),
  z = z1_2,
  type = "surface",
  scene = "scene2"
) %>%
  add_surface(x = unique(xy_grid[, 1]),
              y = unique(xy_grid[, 2]),
              z = matrix(zs1_2, nrow = 101, byrow = TRUE)) %>% 
  add_trace(data = df1_2, x = df1_2[ , 1], y = df1_2[ , 2], z = df1_2[ , 3],
            mode = "markers", type = "scatter3d",
            marker = list(size = 1, color = "black"))%>%
  layout(scene2 = list(camera = list(
    eye = list(
      x = -1.25,
      y = -1.25,
      z = 1.25
    ),
    center = list(x = 0,
                  y = 0,
                  z = 0)
  )))

plotly_figure_s_2_1 <- plot_ly(
  x = unique(xy_grid[, 1]),
  y = unique(xy_grid[, 2]),
  z = z2_1,
  type = "surface", 
  scene = "scene3"
) %>%
  add_surface(x = unique(xy_grid[, 1]),
              y = unique(xy_grid[, 2]),
              z = matrix(zs2_1, nrow = 101, byrow = TRUE)) %>% 
  add_trace(data = df2_1, x = df2_1[ , 1], y = df2_1[ , 2], z = df2_1[ , 3],
            mode = "markers", type = "scatter3d",
            marker = list(size = 1, color = "black"))%>%
  layout(scene3 = list(camera = list(
    eye = list(
      x = -1.25,
      y = -1.25,
      z = 1.25
    ),
    center = list(x = 0,
                  y = 0,
                  z = 0)
  )))

plotly_figure_s_2_2 <- plot_ly(
  x = unique(xy_grid[, 1]),
  y = unique(xy_grid[, 2]),
  z = z2_2,
  type = "surface",
  scene = "scene4"
) %>%
  add_surface(x = unique(xy_grid[, 1]),
              y = unique(xy_grid[, 2]),
              z = matrix(zs2_2, nrow = 101, byrow = TRUE)) %>% 
  add_trace(data = df2_2, x = df2_2[ , 1], y = df2_2[ , 2], z = df2_2[ , 3],
            mode = "markers", type = "scatter3d",
            marker = list(size = 1, color = "black"))%>%
  layout(scene4 = list(camera = list(
    eye = list(
      x = -1.25,
      y = -1.25,
      z = 1.25
    ),
    center = list(x = 0,
                  y = 0,
                  z = 0)
  )))

subplot(plotly_figure_s_1_1, plotly_figure_s_1_2, plotly_figure_s_2_1, plotly_figure_s_2_2, nrows=2)
```